Towards atomistic resolution structure of phosphatidylcholine glycerol backbone and choline headgroup at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-04-07 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
update dehydration discussion after adding Berger DLPC results

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of accurate experimental data, the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulations model 
would automatically resolve the structures giving an interpretation of experimental results, if the model
reproduced the experimental data. 
In the present work we simulate phosphatidylcholine (PC) lipid bilayers with 13 different
atomistic models, and we compare simulations with experiments (in fully hydrated conditions) in terms of C--H bond vector order
parameters for the glycerol backbone and choline headgroups. 
We show that this comparison can be used to judge the atomistic resolution structural accuracy of the models,
thus allowing the usage of molecular dynamics simulations to interpret the structures of biomolecules in 
biologically relevant conditions from NMR experiments. Further, we review previous experimental
data for dehydrated lipid bilayers and cholesterol containing bilayers, and interpretate it with simulations.
The interpretation reveals that the increased choline order parameters indicate P--N vector tilting more parallel to membrane,
and that cholesterol induces only minor changes to the glycerol backbone structure.
%None of the current models is not accurately enough to resolve the structure with experimental accuracy.
%However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
%that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
%would resolve the structure. Despite of the inaccuracy in the fully hydrated
%structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
%is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
%between lipids and cholesterol better than the Berger/H\"oltje model.
This work has been done as an open collaboration by using the \url{nmrlipids.blogspot.fi} as an
communication platform. All the scientific contributions have been done through the blog and
are publicly available. In addition, almost all simulation trajectories and files are made available
in the Zenodo community \url{https://zenodo.org/collection/user-nmrlipids} which has became the most extensive
publicly available collection of molecular dynamics simulation trajectories of lipid bilayers.
\end{abstract}

@@

Lipid bilayer dehydration has been studied also with molecular dynamics simulations~\cite{mashl01,pertsin05,pertsin07,eun09,eun10,schneck12},
typically motivated by the  discussion about the origin of the ``hydration repulsion''~\cite{israelachvili,israelachvili96,sparr11}.
However, the used simulation models are not typically compared to the experimental choline and glycerol backbone
order parameters (except by Mashl et al.~\cite{mashl01}).
In Fig.~\ref{ordPhydr} the glycerol backbone and choline order parameters are shown as a function of hydration level for the CHARMM36, 
MacRog and GAFFlipid models (having the most realistic atomistic resolution structures) together with the Berger based model 
(which is the most used lipid model). 
To elucidate changes the simulation results are shifted to match better with experiments with full hydration.
The magnitude of shifting is written in the y-axis label.
The choline order parameters increase with dehydration in all
models, in qualitative agreement with experiments. 
The measured decrease in both g$_3$ and g$_2$ order parameters with dehydration is reproduced only in CHARMM36.
\todo{This sentence must be rewritten after the up-to-date results for GAFFlipid are in the figure.}

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration in Fig.~\ref{PNangle}.

It should be noted, however, that free energy landscape is not correct in the models which
are not able to reproduce the experimental order parameters. Thus the energetical response
to the dehydration may not be correct even though the order parameter response is qualitatively correct.
This issue may have some influence on the dehydration energetic calculations made with the Berger model~\cite{eun09,schneck12}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../Fig/dehydrationPN.pdf}

@@

The response of the glycerol backbone to dehydration seems to be more subtle than that of the choline headgroup 
as CHARMM36 is the only model that reproduces the decrease in g$_2$ and g$_3$ segments.
\todo{This sentence must be rewritten after the up-to-date results for GAFFlipid are in the figure.}

@@

- The review of previous experimental results revealed that the 
choline order parameters are increasing when the bilayer is dehydrated. Our simulations show that
this can be explained by P--N vector tilting more parallel to the membrane. 
These results strongly supports and complements the idea that the charge induced choline tilting and 
partition can be measured by using the order parameter changes~\cite{ionpaper,scherer89}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of accurate experimental data, the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulations model 
would automatically resolve the structures giving an interpretation of experimental results, if the model
reproduced the experimental data. 
In the present work we simulate phosphatidylcholine (PC) lipid bilayers with 13 different
atomistic models, and we compare simulations with experiments (in fully hydrated conditions) in terms of C--H bond vector order
parameters for the glycerol backbone and choline headgroups. 
We show that this comparison can be used to judge the atomistic resolution structural accuracy of the models,
thus allowing the usage of molecular dynamics simulations to interpret the structures of biomolecules in 
biologically relevant conditions from NMR experiments. Further, we review previous experimental
data for dehydrated lipid bilayers and cholesterol containing bilayers, and interpretate it with simulations.
The interpretation suggest that the increased choline order parameters indicate P--N vector tilting more parallel to membrane,
and that cholesterol induces only minor changes to the glycerol backbone structure.
%None of the current models is not accurately enough to resolve the structure with experimental accuracy.
%However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
%that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
%would resolve the structure. Despite of the inaccuracy in the fully hydrated
%structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
%is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
%between lipids and cholesterol better than the Berger/H\"oltje model.
This work has been done as an open collaboration by using the \url{nmrlipids.blogspot.fi} as an
communication platform. All the scientific contributions have been done through the blog and
are publicly available. In addition, almost all simulation trajectories and files are made available
in the Zenodo community \url{https://zenodo.org/collection/user-nmrlipids} which has became the most extensive
publicly available collection of molecular dynamics simulation trajectories of lipid bilayers.
\end{abstract}

@@

Lipid bilayer dehydration has been studied also with molecular dynamics simulations~\cite{mashl01,pertsin05,pertsin07,eun09,eun10,schneck12},
typically motivated by the  discussion about the origin of the ``hydration repulsion''~\cite{israelachvili,israelachvili96,sparr11}.
However, the used simulation models are not typically compared to the experimental choline and glycerol backbone
order parameters (except by Mashl et al.~\cite{mashl01}).
In Fig.~\ref{ordPhydr} the glycerol backbone and choline order parameters are shown as a function of hydration level for the CHARMM36, 
MacRog and GAFFlipid models (having the most realistic atomistic resolution structures) together with the Berger based model 
(which is the most used lipid model). 
To elucidate changes the simulation results are shifted to match better with experiments with full hydration.
The magnitude of shifting is written in the y-axis label.
The choline order parameter increase with dehydration is seen in all
models despite of some additional fluctuations. Thus the choline order parameter response to dehydration can be
intepreted to be in qualitative agreement with experiments.
The situation is significantly more complicated for the glycerol backbone segments 
and none of the models can be interpreted to qualitatively agree with experiments
for all the segmets.
\todo{DONE}

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration in Fig.~\ref{PNangle}.
However, the amount of increase depends on the model. Especially the DLPC simulations with Berger model
predict significantly stronger P--N vector tilt compared with the other models. The Berger model
has also generally larger P--N vector angles and its choline order parameters are more off from
experiments than other models. Thus the relatively modest tilting with dehydration
predicted by MacRog, CHARMM36 and GAFFlipid is probably more realistic.

It should be also noted that the free energy landscape is not correct in the models which
are not able to reproduce the experimental order parameters. Thus the energetical response
to the dehydration may not be correct even though the order parameter response is qualitatively correct.
This issue may have some influence on the dehydration energetic calculations made with the Berger model~\cite{eun09,schneck12}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../Fig/dehydrationPN.pdf}

@@

The response of the glycerol backbone to dehydration seems to be more subtle than that of the choline headgroup 
and none of the models reproduced the experimental results with the accuracy allowing
the structural interpretation.
\todo{DONE}

@@

- The review of previous experimental results revealed that the 
choline order parameters are increasing when the bilayer is dehydrated. Our simulations suggest that
this can be explained by P--N vector tilting more parallel to the membrane. 
These results strongly supports and complements the idea that the charge induced choline tilting and 
partition can be measured by using the order parameter changes~\cite{ionpaper,scherer89}.
